---
title: "Covid19"
author: "TP"
date: "4/18/2022"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
```

In this project, I take a look at total COVID 19 cases in all the countries in the world. 

# DATA IMPORT
I will start by reading in the data from the four main csv files

```{r get_jhu_data}
## Get current data in the four files
# they all begin the same way
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv", 
                "time_series_covid19_confirmed_US.csv", "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)

```

Let's read in the data and see what we have 

```{r import_data, message=FALSE}
global_cases <- read_csv(urls[1])
global_cases
global_deaths <- read_csv(urls[2])
global_deaths
US_cases <- read_csv(urls[3])
US_cases
US_deaths <- read_csv(urls[4])
US_deaths
```

After looking at global_cases and global_deaths, I would like to tidy those datasets and put each variable (date, cases, deaths) in their own column. 

Also I don't need Lat and Long for the analysis I am planning, so I will get rid of those and rename Region and State to be more R friendly. 

```{r tidy_global_data}
global_cases <- global_cases %>% 
  pivot_longer(cols = -c('Province/State', 
                         'Country/Region', Lat, Long),
               names_to = "date", 
               values_to = "cases") %>%
  select(-c(Lat, Long))

global_cases

global_deaths <- global_deaths %>% 
  pivot_longer(cols = -c('Province/State', 
                         'Country/Region', Lat, Long),
               names_to = "date", 
               values_to = "deaths") %>%
  select(-c(Lat, Long))

global_deaths
                    
```

Combining the cases into deaths per date into one variable we will call global
Fixing date presentation
```{r tidy_2}
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region', 
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

global
# Filter out and keep only where the cases are positive
global <- global %>% filter(cases > 0)
global
```

Add a population data and a variable called Combined_key that combines these two things (province_state and the country_region together so that I'll have a similar sort of dataset )
```{r global_add_population}
global <- global %>%
  unite("Combined_Key", 
        c(Province_State, Country_Region),
        sep = ", ", 
        na.rm = TRUE,
        remove = FALSE)

global

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
uid 

global <- global %>%
  left_join(uid, by=c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
global
```

# VISUALIZING
```{r visualizing}
global_by_country <- global %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
global_by_country

country = "Australia"

global_by_country %>% 
  filter(Country_Region == country) %>%
  filter(cases > 0 ) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color="deaths")) +
  geom_point(aes(y = deaths, color="deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID19 in ", country), y=NULL)


global_totals <- global_by_country %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
global_totals

tail(global_totals) 


# VISUALIZE
global_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom", 
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in the World", y = NULL)

max(global_totals$date)
max(global_totals$deaths)
```


# ANALYZING

``` {r analyzing}
global_by_country <- global_by_country %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

global_totals <- global_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(global_totals)

tail(global_totals %>% select(new_cases, new_deaths, everything()))

global_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 new cases/deaths in the World", y=NULL)

global_by_country %>%
  filter(Country_Region == country) %>%
  filter(cases > 0 ) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color="new_deaths")) +
  geom_point(aes(y = new_deaths, color="new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID19 in ", country), y=NULL)

# WHAT ARE THE WORST AND BEST COUNTRIES? SHOULD WE LOOK AT THE TOTAL CASES OR THE DEATH RATES PER 1000 PEOPLE?

global_country_totals <- global_by_country %>%
  group_by(Country_Region) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases/population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

global_country_totals %>%
  slice_min(deaths_per_thou, n = 10) #states has the smallest number of deaths_per_thousand
# 10 countries has the lowest COVID deaths
global_country_totals

global_country_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
# 10 countries has the highest COVID deaths
global_country_totals
```


# MODELING

```{r modeling}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = global_country_totals)
summary(mod)

global_country_totals %>% slice_min(cases_per_thou)

global_country_totals %>% slice_max(cases_per_thou)

x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thou = x_grid)
global_country_totals %>% mutate(pred = predict(mod))

global_tot_w_pred <- global_country_totals %>% mutate(pred = predict(mod))

global_tot_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")

```

# BIAS
There can be a bias in how the data has been collected. For example, people who go to hospital or report to workplace are more likely to be included in the data. On the other hand, people who live in the countryside, elders who stay at home almost all the time are less likely to be recorded. It also depends on which testing sides that are used to collect the data and it's different all over countries. The correct case number may be much higher than this, depending on how correctly each country records their cases. 